Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-05-12 manuscript/main.qmd
update results, reformat suppl. material

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
editor: source
---

@@

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34°N--32 degS), with minimum sea surface temperature (\~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that can be confidently assigned to the EECO. We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34 deg N). Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6 degC) and *Rhizophora* (20.7 degC) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*). However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 degC and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF). We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.

We also used our model to estimate the modern sea surface temperature gradient, using annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131)

To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value. We thus substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations. The re-sampled, simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0 deg latitude) and at the poles (90 deg absolute latitude). Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), this results in only slightly higher estimates of the gradient than when comparing e.g. the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [e.g. @Evans2018].

@@

Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$ where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.

@@

@@

# modern R2
R2_modern <- readRDS("results/modern/modern_model_empirical_R2.rds")

@@

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination (*R*^2^ value) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown. Each gradient is depicted in absolute latitude.
:::

```{=html}
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
```
\[Fig 3 - Modern gradient\]

@@

Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model. The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature. Grey points depict the individual cell values of the Bio-ORACLE grid.
:::

The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"` degC when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.

@@

The modeled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the EECO, as opposed to `r modern_gradient$"50%"` degC for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6 deg), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`).

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC higher than the modern. A model run excluding the ecological proxies results in a GMsST that is higher by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`) degC. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

>>>>

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
header-includes:
  - \usepackage{newunicodechar}
  - \newcommand{\$^\circ$C}{\textdegree C}
bibliography: references.bib
csl: climate-of-the-past.csl
editor: source
---

@@

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@Royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature (\~18$^\circ$C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 $^\circ$N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that can be confidently assigned to the EECO. We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N). Subsequently, we use statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*). However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05$^\circ$C and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.

2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7$^\circ$C (lower temperature limit of *Rhizophora*). As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities as mangroves are also widely distributed throughout tropical regions. Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF). We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions. To assess how well the modeled gradients reconstructed from limited sampling fit the idealised gradients, we calculated an R^2 for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the R^2 based on these values. We report the median, and 2.5th and 97.5th percentile of the resulting R^2 measures.

To test whether out model can accurately model the the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131). The R^2 for the modern gradient was calculated as described above, comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value, making the hierarchical structure superfluous. We thus substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modeled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).

@@

The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude). Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), this results in only slightly higher estimates of the gradient than when comparing e.g. the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [e.g. @Evans2018].

@@

Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins. The weights were set to the proportion of global surface area in each latitudinal bin, i.e. decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$.

@@

modern_gradient_empirical <- readRDS("results/modern/modern_empirical_gradient.rds")

@@

# modern R2
R2_modern <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")

@@

## Model validation and modern gradient

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination (*R*^2) for estimating goodness of fit between the simulated and modeled gradient. The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown. Each gradient is depicted in absolute latitude.
:::

```{=html}
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
```

@@

Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model. The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperatures. Grey points depict the individual cell values of the Bio-ORACLE grid.
:::

Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to "super greenhouse" scenarios (Fig. 2). Random latitudinal sampling results in highly accurate reconstructions already at a sample size of 10 for the icehouse scenarios (95 % CI of R^2 > 0.9). Greenhouse scenarios require more random samples to accurately predict high-latitude temperatures. This is because in the absence of high-latitude samples, the modeled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient. A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of R^2 > 0.95).

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modeled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C. The modern, global men sea surface temperature based on our modeled gradient is `r motemp_global$"50%"`$^\circ$C.

<!-- REMOVED
Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles. When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`)$^\circ$C. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`$^\circ$C (95 % CI).

@@

Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`$^\circ$C when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`)$^\circ$C. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`$^\circ$C.
-->

@@

The modeled Eocene temperature gradient is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`)$^\circ$C. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`)$^\circ$C, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern. A model run excluding the ecological proxies increases the GMsST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C. The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig. S3).
